Sparse Function-space Representation of Neural Networks

### 2023-08-31 workshops/icml-duality/main.tex

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits `dual' parameters which are the building blocks of a GP~\citep{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \citep{immer2021scalable}, this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs.
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\citep{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network; see \cref{fig:teaser} for an example. %Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach for conditioning on new data without needing to retrain the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

\textbf{Linearization gives rise to a Gaussian process}
Our goal is to capture the distribution over the NN model functions through their first two moments. The first two moments characterize a GP with a mean function $\mu(\cdot)$ and a covariance function (kernel) $\kappa(\cdot,\cdot)$.
Recent work \citep{khan2019approximate,maddox2021fast} has shown that linearizing approximations in the weight space lead to function-space equivalent approximations.
As Gaussian distributions remain tractable under linear transformations, a linear function in terms of parameters can be converted from the weight space to the function space \citep[see Ch.~2.1 in ][]{rasmussen2006gaussian} as follows:
%
\begin{align} \label{eq:weight_func}
 &f_\weights(\vx) \approx
%g_\weights(\mathbf{x}) =
 \phi^\top\!(\vx) \, \vw \quad\implies \nonumber \\
 &\mu(\vx) = 0 \quad \text{and} \quad \kappa(\vx, \vx') = \frac{1}{\delta} \phi^\T\!(\vx) \, \phi(\vx').
\end{align}
% The posterior structure directly relates to the optimization loss around the MAP weights $\vw^*$.
A common approach is to approximate the correlation structure of a distribution centred at the MAP estimate as done in the Laplace-GGN~\citep{khan2019approximate, daxberger2021laplace, maddox2021fast}.
The Laplace-GGN takes the MAP solution and approximates the Hessian of the loss function
with the GGN.
Following the work of \citet{khan2019approximate}, we can view this as a linear approximation of the NN, as $f_{\weights^*}(\vx) \approx
%g_{\weights}(\vx) =
\Jac{\weights^*}{\vx} \, \weights$, where $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{1 \times P}$ is the Jacobian at the MAP.
For notational conciseness, we % \cref{eq-laplace-approx-function-space},
restrict our notatin to a single function output. The extension to multiple outputs is straightforward.

Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior over $\vw$.
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space,
\begin{equation}
\label{eq-laplace-approx-function-space}
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'},
\end{equation}
where the kernel is the so-called Neural Tangent Kernel \citep[NTK,][]{jacot2018neural}.
We can combine this kernel function with the data set $\dataset$ to construct the posterior.
%
\citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the NN to form a Bayesian GLM model. Both approaches adjust the GP posterior mean function to ensure the predictions are from the NN $f_{\vw^*}(\vx)$.
In contrast, our method makes predictions directly from the GP and avoids such adjustments.
In the next section, we present our method for converting a trained NN to a sparse functional representation.
% on the non-sparse approximation.
% of our
% based on the non-sparse approximation.
% we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

@@

\paragraph{Dual parameters from NN}
Given that we use a Laplace approximation of the NN, we remove the expectation over the posterior \citep[see Ch.~3.4.1 in][for derivation]{rasmussen2006gaussian} and our dual parameters are given by,
%
\begin{align}
  \label{eq:dual_param_laplace}
  \hat{\alpha}_i &\coloneqq \nabla_{f}\log p(y_i \mid f) |_{f=f_i} , \\
  %\quad \text{and} \quad
  \hat{\beta}_i &\coloneqq - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i}.
\end{align}
%
Substituting \cref{eq:dual_param_laplace} into \cref{eq:gp_pred}, we obtain our GP based on the converged NN. Again, this is similar to what was derived in \citet{immer2021improving} for the posterior variance function.
However, they use the NN's prediction $f_{\vw^*}(\cdot)$ for the posterior mean and do not emphasize the significance of the dual parameters.
The problem with the predictions in \cref{eq:gp_pred} is that they incur a cost of $O(N^3)$, which limits its application on large data sets.

>>>>

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits `dual' parameters which are the building blocks of a GP~\citep{csato2002sparse, adam2021dual, chang2023memory} . In contrast to previous work that utilizes subsets of training data \citep{immer2021scalable}, this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs.
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\citep{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network; see \cref{fig:teaser} for an example. %Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach for conditioning on new data without needing to retrain the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

\textbf{Linearization gives rise to a Gaussian process}
Our goal is to capture the distribution over the NN model functions through their first two moments. The first two moments characterize a GP with a mean function $\mu(\cdot)$ and a covariance function (kernel) $\kappa(\cdot,\cdot)$.
Recent work \citep{khan2019approximate,maddox2021fast} has shown that linearizing approximations in the weight space lead to function-space equivalent approximations.
As Gaussian distributions remain tractable under linear transformations, a linear function in terms of parameters can be converted from the weight space to the function space \citep[see Ch.~2.1 in ][]{rasmussen2006gaussian} as follows:
%
\begin{align} \label{eq:weight_func}
 &f_\weights(\vx) \approx
%g_\weights(\mathbf{x}) =
 \phi^\top\!(\vx) \, \vw \quad\implies \nonumber \\
 &\mu(\vx) = 0 \quad \text{and} \quad \kappa(\vx, \vx') = \frac{1}{\delta} \phi^\T\!(\vx) \, \phi(\vx').
\end{align}
% The posterior structure directly relates to the optimization loss around the MAP weights $\vw^*$.
A common approach is to approximate the correlation structure of a distribution centred at the MAP estimate as done in the Laplace-GGN~\citep{khan2019approximate, daxberger2021laplace, maddox2021fast}.
The Laplace-GGN takes the MAP solution and approximates the Hessian of the loss function
with the GGN, we can view this as a linear approximation of the NN, as $f_{\weights^*}(\vx) \approx
%g_{\weights}(\vx) =
\Jac{\weights^*}{\vx} \, \weights$, where $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{1 \times P}$ is the Jacobian at the MAP.
For notational conciseness, we % \cref{eq-laplace-approx-function-space},
restrict our notatin to a single function output. The extension to multiple outputs is straightforward.

Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior over $\vw$.
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space,
\begin{equation}
\label{eq-laplace-approx-function-space}
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'},
\end{equation}
where the kernel is the so-called Neural Tangent Kernel \citep[NTK,][]{jacot2018neural}.
We can combine this kernel function with the data set $\dataset$ to construct the posterior.
%
We highlight that  \citet{immer2021improving}  adjust the GP posterior mean function to match the NN predictions $f_{\vw^*}(\cdot)$. In contrast, our method makes predictions directly from the GP and avoids such adjustments. In the next section, we present our method for converting a trained NN to a sparse functional representation.

%took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the NN to form a Bayesian GLM model. Both approaches adjust the GP posterior mean function to ensure the predictions are from the NN $f_{\vw^*}(\vx)$.
%In contrast, our method makes predictions directly from the GP and avoids such adjustments.
%In the next section, we present our method for converting a trained NN to a sparse functional representation.
% on the non-sparse approximation.
% of our
% based on the non-sparse approximation.
% we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

@@

\paragraph{Dual parameters from NN}
Given that we use a Laplace approximation of the NN, we remove the expectation over the posterior \citep[see Ch.~3.4.1 in][for derivation]{rasmussen2006gaussian} and our dual parameters are given by,
%
\begin{equation}
\begin{split}
  \label{eq:dual_param_laplace}
  \hat{\alpha}_i &\coloneqq \nabla_{f}\log p(y_i \mid f) |_{f=f_i} , \\
  %\quad \text{and} \quad
  \hat{\beta}_i &\coloneqq - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i}.
  \end{split}
\end{equation}
%
Substituting \cref{eq:dual_param_laplace} into \cref{eq:gp_pred}, we obtain our GP based on the converged NN. The problem with forming predictions using \cref{eq:gp_pred} is that they cost $O(N^3)$, which limits a full GP model to application on large data sets.
